Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-09 MScDiplom_Lemenkova.tex
Written 'Discussion', added Appendix 2 (illustrations of some RS concepts), Appendix 10 (snapshots of the working process)

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
\usepackage[english]{babel}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
\usepackage{pslatex}
%\usepackage{times}
%\usepackage{psnfss}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue}

@@

@@

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}\index[people]{Toxopeus, A.G.}  and Mr. \textit{Valentijn Venus}\index[people]{Venus, V.} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. 
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis}\index[people]{Limberakis, Petros} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus}\index[people]{Venus, V.} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using GDAL technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}\index[people]{Gege P.}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}\index[people]{Lemenkov, P.}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}\index[people]{Nieuwenhuis, W.}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie}\index[people]{de Bie C.A.J.M.} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr. \textit{Dingtian Yang}\index[people]{Yang D.} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

The seagrasses, a unique group of aquatic plants, create complex, extremely diversified and productive ecological systems in the littoral coastal zones. The only flowering plant in the world that is able to live completely submerged, seagrasses play vital role in the marine ecosystems of the World Ocean. Seagrasses are the most important component in the environmental food chain of the coastal ecosystems, being a vital food source for various marine species (e.g. fish, dugongs, turtles, swans), and a producer of organic matter, which is the very basis of the food web.\\
The \textit{P.oceanica} seagrass is an endemic for the Mediterranean region, and a main species in the marine coastal environment of Greece. Meadows of \textit{P.oceanica} are subjected to the human activities, because they occur in coastal areas, where they are affected both by anthropogenic and by climatic and environmental factors. Nowadays \textit{P.oceanica} is in the alarming state of regression, because of the deterioration of the environment in the Mediterranean Sea. Due to these reasons, \textit{P.oceanica} is a protected species since 1988 in some European countries (France). Monitoring \textit{P.oceanica} is therefore an important contribution to the saving and protecting the environment of Mediterranean region.\\
The current MSc thesis focuses on the monitoring of seagrass \textit{P.oceanica} along the northern coasts of Crete Island, Greece, and investigates the application of the remote sensing techniques for the seagrass mapping. 
This research was articulated in two parts, where the first one involves an ecological approach to the seagrass distribution in various regions around the globe and the experience of seagrass monitoring nowadays. The second part of this work has technical character and investigates the application of the remote sensing techniques towards seagrass mapping. It, furthermore, focuses on the optical properties of the \textit{P.oceanica} and other seafloor cover types, and studies distinguishability of various seafloor cover types. Studies of the optical characteristics of separate seafloor cover types were made with purpose to clarify, whether their spectral properties change with varying environmental conditions. \\
Special attention has been drawn on the role of environmental factors on the distribution of \textit{P.oceanica} along the coasts of Crete, and in particular, how the optical properties of the seafloor cover types, i.e. spectral reflectance, are being changed under varying external conditions, e.g. water column, amount of suspended particles and sediments in the seawater, and water temperature. For this purpose we studied differences in the spectral reflectance of \textit{P.oceanica} and other bottom cover types at three distinct depths. The diverse spectral values entail variations in optical properties of the seafloor cover types at changing environmental conditions. We applied \ac{WASI} simulation techniques for the modelling of the optical parameters of various seafloor cover types by various spaceborne imaging spectrometers (\ac{MERIS}, \ac{SeaWiFS}, \ac{CZCS} and \ac{MODIS}), in order to understand their suitability and possible limitations for the seagrass mapping.\\
Fieldwork research sites were presented by separate locations on the northern coast of Crete region (Ligaria, Agia Pelagia, Xerocampos). The additional measurements of the reflectance spectra of the seawater with and without sediments have been made in aquarium tank in 2009 by means of Trios-RAMSES spectroradiometer. Parallel to the collection of spectra signatures, we captured the imagery for the seagrass mapping, which consists of the aerial images from the Google Earth website and the satellite Landsat TM and ETM+ scenes. 
\pagebreak

@@

@@

@@

	\vspace{1ex}
	\textbf{Main Tools}:
	\begin{itemize}
		\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
		\item \href{http://www.ilwis.org/}{ILWIS}, \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and classification
		\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water 					properties under changing environment
		\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and \ac{GPS} for data capture during the fieldwork
		\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
		\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
		\item \href{http://www.gnuplot.info/}{GNUplot} for plotting graphs
		\item \href{http://gretl.sourceforge.net/}{Gretl: GNU Regression, Econometrics and Time-series Library} for statistic analysis
		\item \href{http://www.openoffice.org/}{Open Office}: document typesetting, data pre-processing and preliminary statistical evaluation		
		\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)}: statistical data edition and analysis

		\item \href{http://staruml.sourceforge.net/en/}{StarUML}, \href{http://live.gnome.org/Dia}{Dia}, \href{http://argouml.tigris.org/}						{ArgoUML}: \ac{UML} diagram tool for creating flowcharts of work algorithms
	\end{itemize}

@@

The spectral reflectance, distinct for each seafloor cover type, enables their discrimination on the image. The pre-processing
of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).
\pagebreak

@@

\subsection[Measuring water optical properties]{Measuring water optical properties: \\hyperspectral radiometers}
Among various research approaches for the seagrass mapping the most popular methods are based on application of the remote sensing techniques,
 enabling to detect and discriminate the substratum and various vegetation types growing on the seafloor.
The application of the remote sensing data for seagrass mapping is based on our knowledge of the
spectral reflectance properties of the target objects, and using it for the classification of these objects
on the image. In case of seagrasses it is spectral reflectance of the seafloor cover types, which can be
analyzed using measurements of optical properties of sea water: radiance and irradiance.
Optical remote sensing methods can get through the clear waters to approximately 15–30 m \cite{Mumby04}\label{Mumby04}.
When sunlight enters the waters and goes down into the water column, parts of the
electromagnetic energy are absorbed and scattered, which is determined by the optical and physical
properties of the water, e.g. concentration of suspended particles, chlorofill, colored dissolved
organic matter (gelbstoff) that make up the water content \cite{Ackleson2003}\label{Ackleson, 2003}. Besides, light is strongly
dependent on wavelengths, i.e. it is greater in blue wavelengths (400 nm) than in others.

@@

\subsection[Radiative Transfer Models (RTM)]{Radiative Transfer Models (RTM) for the simulation of water optical properties: a brief review of existing tools}

@@

\subsection[Google Earth aerial imagery]{Google Earth aerial imagery for the seagrass mapping}
Apart from the satellite imagery, the aerial photographs from the Google Earth provide
a powerful tool for seagrass mapping, because they are important, reliable, detailed
and up-to-date source of imagery.
Perhaps, the clearest advantage of the Google Earth imagery is its high resolution (15 m in land areas and lower in the oceans).
Obtained from the airborne platforms, Google Earth images have general spatial
resolution of several meters (though varying in different areas), which allows very
detailed habitat and seafloor types discrimination, comparing with images
acquired from the space-borne satellite platforms.\\
The spatial coverage of the Google imagery is lesser comparing to spaceborne images, but this can be solved as
well: while in general providing smaller area coverage than satellites images, the Google Earth
images can be stitched to the composite maps of the acceptable spatial extent, using Google grabbing
process which allows multiple overlapping of single images over the flight paths, and generates
mosaics.

@@

Besides spatial structure of the seagrass meadows, there are strong and complex patterns of depth zoning, specific to individual
seagrass species, but such detailed classification goes beyond the scope of the current MSc research.
Unique feature of seagrass meadows, most characteristic for shallow areas, is dynamics of their
landscapes. Homogeneous, continuous seagrass meadows can reach in size up to 40 m2 (Fig.\ref{fig:38} a),
however they are often interrupted by gaps and channels of open spaces generated by the complex,
disturbing sedimentation processes, turbulence and turbidity regime of waters and landscape dynamics,
which form gaps and channels within meadows, making them “patchy-looking”, i.e. diversified by
separate mattes (Fig.\ref{fig:38} b). The traditional definition for gap in vegetation cover is “disturbance generated
openings in either floral or faunal cover” \cite{Connell78}\label{Connell78}.
Formation and increase of these
gaps within the mosaic of seagrass meadows is caused by different reasons. The most probable
drivers for the process of gaps formation within seagrass meadows are removal of interior vegetation,
differential growth of seagrass meadows and increased sedimentation. Thus, storms lead to severe
deposition of sediments, burying parts of seagrass meadow, the same effect has movements and
deposition of sediments during and after floods\cite{Bell99}\label{Bell99}.
Increased nutrient sedimentation, especially phosphorus, were explored \cite{Jensen01}\label{Jensen01} as a
potential mechanism for increasing patch dynamics and morphological plasticity within seagrass
meadows. Finally, increasing the degree of fragmentation of the landscapes of 	extit{P.oceanica} meadows
can be caused by the invasion of alien species, such as Cymodocea nodosa, Caulerpa prolifera,
Caulerpa taxifolia. Invaders are in general strong colonizers comparing to native \textit{P.oceanica}: they
occupy much greater habitat space within the regressed meadows of stressed native seagrass
\cite{Montefalcone10}\label{Montefalcone10}. However, on the northern coasts of Crete the only dominating seagrass
species is \textit{P.oceanica}.
Morphological differences in scale of seagrass landscape formations, discussed above, cause need for
the different-scale mapping. Therefore, the investigation of the seagrass meadows at different levels is
performed using underwater videometric measurements, aerial and satellite imagery.

@@

An approach of the seagrass spectral analysis, monitoring and mapping has been taken in this work, which integrates various research techniques and tools, combining remote sensing methods of spectral analysis of the seafloor cover types, and knowledge of the ecology of \textit{P.oceanica}, with the aim to develop a method of seagrass spectral optical discrimination for the seagrass mapping based on imagery classification.\vspace{1ex}\\
The ecological variables, specific to the field environmental conditions, were factored into the \ac{WASI}-based simulation models. Through the \ac{WASI} simulation process, imitating spectral properties of \textit{P.oceanica} and carbonate sand for various broadband and narrowband sensors, models were created that accounted for not only atmospheric conditions (i.e. sun zenith angle), but also height of water column, thus approaching it to the Mediterranean conditions, and chemical content of the seawater (i.e. amount of suspended particles, Gelbstoff, etc), which results in models of optical properties of ``seawater with sediments" and ``seawater without sediments".\vspace{1ex}\\

>>>>

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
\usepackage[english]{babel}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
\usepackage{pslatex}
%\usepackage{times}
%\usepackage{psnfss}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue,pdfkeywords={P.oceanica, seagrass, spectral reflectance, remote sensing, WASI, RTM, Google Earth, mapping, Crete}}

@@

\clearpage
\section*{Dedication}
	\markright{Dedication}
	\addcontentsline{toc}{section}{Dedication}
%\begin{flushright}
\newenvironment{dedication}
    {\vspace{16em}\begin{quotation}\begin{center}\begin{em}}
    {\par\end{em}\end{center}\end{quotation}}
\begin{dedication}
\large{\textsc{To my elder brother and best friend, \textbf{Peter Lemenkov}\index[people]{Lemenkov, P.}}}
\end{dedication}
%\end{flushright}
\pagebreak

@@

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}\index[people]{Toxopeus, A.G.}  and Mr. \textit{Valentijn Venus}\index[people]{Venus, V.} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. 
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, Department of Zoology of the University of Crete, especially to Dr.\textit{Petros Lymberakis}\index[people]{Lymberakis, P.} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus}\index[people]{Venus, V.} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using GDAL technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}\index[people]{Gege P.}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}\index[people]{Lemenkov, P.}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}\index[people]{Nieuwenhuis, W.}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie}\index[people]{de Bie C.A.J.M.} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr. \textit{Dingtian Yang}\index[people]{Yang D.} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

The seagrasses, a unique group of aquatic plants, create complex, extremely diversified and productive ecological systems in the littoral coastal zones. The only flowering plant in the world that is able to live completely submerged, seagrasses play vital role in the marine ecosystems of the World Ocean. Seagrasses are the most important component in the environmental food chain of the coastal ecosystems, being a vital food source for various marine species (e.g. fish, dugongs, turtles, swans), and a producer of organic matter, which is the very basis of the food web.\\
The \textit{P.oceanica} seagrass is an endemic for the Mediterranean region, and a main species in the marine coastal environment of Greece. Meadows of \textit{P.oceanica} are subjected to the human activities, because they occur in coastal areas, where they are affected both by anthropogenic and by climatic and environmental factors. Nowadays \textit{P.oceanica} is in the alarming state of regression, because of the deterioration of the environment in the Mediterranean Sea. Due to these reasons, \textit{P.oceanica} is a protected species since 1988 in some European countries (France). Monitoring \textit{P.oceanica} is therefore an important contribution to the saving and protecting the environment of Mediterranean region.\\
The current MSc thesis focuses on the monitoring of seagrass \textit{P.oceanica} along the northern coasts of Crete Island, Greece, and investigates the application of the remote sensing techniques for the seagrass mapping. 
This research was articulated in two parts, where the first one involves an ecological approach to the seagrass distribution in various regions around the globe and the experience of seagrass monitoring nowadays. The second part of this work has technical character and investigates the application of the remote sensing techniques towards seagrass mapping. It, furthermore, focuses on the optical properties of the \textit{P.oceanica} and other seafloor cover types, and studies distinguishability of various seafloor cover types. Studies of the optical characteristics of separate seafloor cover types were made with purpose to clarify, whether their spectral properties change with varying environmental conditions. \\
Special attention has been drawn on the role of environmental factors on the distribution of \textit{P.oceanica} along the coasts of Crete, and in particular, how the optical properties of the seafloor cover types, i.e. spectral reflectance, are being changed under varying external conditions, e.g. water column, amount of suspended particles and sediments in the seawater, and water temperature. For this purpose we studied differences in the spectral reflectance of \textit{P.oceanica} and other bottom cover types at three distinct depths. The diverse spectral values entail variations in optical properties of the seafloor cover types at changing environmental conditions. We applied \ac{WASI} simulation techniques for the modelling of the optical parameters of various seafloor cover types by various spaceborne imaging spectrometers (\ac{MERIS}, \ac{SeaWiFS}, \ac{CZCS} and \ac{MODIS}), in order to understand their suitability and possible limitations for the seagrass mapping.\\
Fieldwork research sites were presented by separate locations on the northern coast of Crete region (Ligaria, Agia Pelagia, Xerocampos). The additional measurements of the reflectance spectra of the seawater with and without sediments have been made in aquarium tank in 2009 by means of Trios-RAMSES spectroradiometer. Parallel to the collection of spectra signatures, we captured the imagery for the seagrass mapping, which consists of the aerial images from the Google Earth website and the satellite Landsat TM and ETM+ scenes.

\section*{Keywords}
\markright{Keywords}
\addcontentsline{toc}{section}{Keywords}
\textit{P.oceanica}, seagrass monitoring, spectral reflectance, remote sensing, WASI, Google Earth, Crete

@@

	\acro{CDOM}{\href{http://en.wikipedia.org/wiki/CDOM}{Colored Dissolved Organic Matter}}

@@

	\acro{FOV}{\href{http://en.wikipedia.org/wiki/FOV}{Field Of View}}

@@

	\vspace{1ex}
	\textbf{Main Tools}:
	\begin{itemize}
		\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
		\item \href{http://www.ilwis.org/}{ILWIS}, \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and classification
		\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water 					properties under changing environment
		\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and \ac{GPS} for data capture during the fieldwork
		\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
		\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
		\item \href{http://www.gnuplot.info/}{GNUplot} for plotting graphs
		\item \href{http://gretl.sourceforge.net/}{Gretl: GNU Regression, Econometrics and Time-series Library} for statistic analysis
		\item \href{http://www.openoffice.org/}{Open Office}: document typesetting, data pre-processing and preliminary statistical evaluation		
		\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)}: statistical data edition and analysis 
		\item \href{http://fwtools.maptools.org/}{FWTools}: for operating images formats	
		\item \href{http://staruml.sourceforge.net/en/}{StarUML}, \href{http://live.gnome.org/Dia}{Dia}, \href{http://argouml.tigris.org/}						{ArgoUML}: \ac{UML} diagram tool for creating flowcharts of work algorithms
	\end{itemize}

@@

The spectral reflectance, distinct for each seafloor cover type, enables their discrimination on the image. The pre-processing
of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).

\subsection{Research assumptions}\index[concepts]{Research!assumptions}
The wind roughening the water surface causes sun glitter, and the values spectral reflectance may contain disturbances and noisy if the solar zenith angle is too high, say approaching 60-70 degrees. Therefore, the measurements should be made at midday (noon) with as low zenith angle (Fig.\ref{fig:A.2-1})\label{zenith} as possible.
\pagebreak

@@

\subsection[Measuring water optical properties]{Measuring water optical properties: \\hyperspectral radiometers}\label{sec:2-2}
Among various research approaches for the seagrass mapping the most popular methods are based on application of the remote sensing techniques,
 enabling to detect and discriminate the substratum and various vegetation types growing on the seafloor.
The application of the remote sensing data for seagrass mapping is based on our knowledge of the
spectral reflectance properties of the target objects, and using it for the classification of these objects
on the image. In case of seagrasses it is spectral reflectance of the seafloor cover types, which can be
analyzed using measurements of optical properties of sea water: radiance and irradiance.
Optical remote sensing methods can get through the clear waters to approximately 15–30 m \cite{Mumby04}\label{Mumby04}.
When sunlight enters the waters and goes down into the water column, parts of the
electromagnetic energy are absorbed and scattered, which is determined by the optical and physical
properties of the water, e.g. concentration of suspended particles, chlorofill, colored dissolved
organic matter (gelbstoff) that make up the water content \cite{Ackleson2003}\label{Ackleson, 2003}. Besides, light is strongly
dependent on wavelengths, i.e. it is greater in blue wavelengths (400 nm) than in others.

@@

\subsection[Radiative Transfer Models (RTM)]{Radiative Transfer Models (RTM) for the simulation of water optical properties: a brief review of existing tools}\label{sec:2-3}

@@

\subsection[Google Earth aerial imagery]{Google Earth aerial imagery for the seagrass mapping}
Apart from the satellite imagery, the aerial photographs from the Google Earth provide
a powerful tool for seagrass mapping, because they are important, reliable, detailed
and up-to-date source of imagery.
Perhaps, the clearest advantage of the Google Earth imagery is its high resolution (15 m in land areas and lower in the oceans).
Obtained from the airborne platforms, Google Earth images have general spatial
resolution of several meters (though varying in different areas), which allows very
detailed habitat and seafloor types discrimination, comparing with images
acquired from the space-borne satellite platforms.\\
The spatial coverage of the Google imagery is lesser comparing to spaceborne images, but this can be solved as
well: while in general providing smaller area coverage than satellites images, the Google Earth
images can be stitched to the composite maps of the acceptable spatial extent, using script written on Python for the \textit{Google Earth} aerial imagery grabbing process (Fig.\ref{fig:A.1} and Fig.\ref{fig:A.2})\label{gdal} which allows multiple overlapping of single images over the flight paths into one swath, and generates imagery mosaics.

@@

Besides spatial structure of the seagrass meadows, there are strong and complex patterns of depth zoning, specific to individual
seagrass species, but such detailed classification goes beyond the scope of the current MSc research.
Unique feature of seagrass meadows, most characteristic for shallow areas, is dynamics of their
landscapes. Homogeneous, continuous seagrass meadows can reach in size up to 40 m2 (Fig.\ref{fig:38} a),
however they are often interrupted by gaps and channels of open spaces generated by the complex,
disturbing sedimentation processes, turbulence and turbidity regime of waters and landscape dynamics,
which form gaps and channels within meadows, making them “patchy-looking”, i.e. diversified by
separate mattes (Fig.\ref{fig:38} b). The traditional definition for gap in vegetation cover is “disturbance generated
openings in either floral or faunal cover” \cite{Connell78}\label{Connell78}.
Formation and increase of these
gaps within the mosaic of seagrass meadows is caused by different reasons. The most probable
drivers for the process of gaps formation within seagrass meadows are removal of interior vegetation,
differential growth of seagrass meadows and increased sedimentation. Thus, storms lead to severe
deposition of sediments, burying parts of seagrass meadow, the same effect has movements and
deposition of sediments during and after floods\cite{Bell99}\label{Bell99}.
Increased nutrient sedimentation, especially phosphorus, were explored \cite{Jensen01}\label{Jensen01} as a
potential mechanism for increasing patch dynamics and morphological plasticity within seagrass
meadows. Finally, increasing the degree of fragmentation of the landscapes of 	extit{P.oceanica} meadows
can be caused by the invasion of alien species, such as \textit{Cymodocea nodosa, Caulerpa prolifera,
Caulerpa taxifolia}. Invaders are in general strong colonizers comparing to native \textit{P.oceanica}: they
occupy much greater habitat space within the regressed meadows of stressed native seagrass
\cite{Montefalcone10}\label{Montefalcone10}. However, on the northern coasts of Crete the only dominating seagrass
species is \textit{P.oceanica}.
Morphological differences in scale of seagrass landscape formations, discussed above, cause need for
the different-scale mapping. Therefore, the investigation of the seagrass meadows at different levels is
performed using underwater videometric measurements, aerial and satellite imagery.

@@

An approach of the seagrass spectral analysis, monitoring and mapping has been taken in this work, which integrates various research techniques and tools, combining remote sensing methods of spectral analysis of the seafloor cover types, and knowledge of the ecology of \textit{P.oceanica}, with the aim to develop a method of seagrass spectral optical discrimination for the seagrass mapping based on the aerial imagery classification.\vspace{1ex}\\
In \textit{Chapter 1:} \nameref{sec:1} we discusses main objective of this MSc thesis, which was to study possibilities of seagrass mapping, based on the application of the remote sensing measurement of the seawater optical properties using hyperspectral radiometers. 
The relationship between the optical properties (spectral reflectance) of the seafloor cover types and hydrological parameters of the environment has been studied in order to analyse limitations and capabilities of broadband and narrowband sensors under the conditions of altering environmental parameters. For the retrieval of hydrological parameters - seawater radiance, irradiance and spectral reflectance of various seafloor cover types, - spectral optical field measurements were carried out in 2009 at the testing sites in Agia Pelagia, Heraklion and Ligaria beach, Crete Island using \ac{Trios-RAMSES} spectroradiometers.\\
Further in \textit{Chapter 1:} 1.8 Research assumptions, we assume the constant values of the optical properties of the seawater, phytoplankton, total amount of suspended particles and solids, atmospheric conditions, as well as \ac{CDOM}, which have been set up in modelling part of this work, during \ac{WASI} simulations of various remote sensors.\vspace{1ex}\\
